Heuristics and Really Hard Instances for Subgraph Isomorphism Problems

### 2016-01-27 paper/paper.tex
Muck around some more

\author{First Author \thanks{This work was supported by a small pile of money, but not as much as we
    would have liked.} \and Second Author
    \and Third Author \\
Some Instituion, Somewhere, Some Country \\
anonymous@example.com}

@@

    \caption{There is a non-induced isomorphism from the pattern graph, in the center, to the target
    on the right, mapping vertex 1 to 1, 2 to 3, 3 to 5 and 4 to 6. This is not an induced
    isomorphism, since there is an edge between 1 and 5 in the target but not between 1 and 3 in the
    pattern. The mapping from the pattern to the (same) target on the left, sending 1 to 2, 2 to 6,
    3 to 5 and 4 to 3, is both non-induced and induced.}
    \label{figure:sip}
\end{figure}

@@

Here we present and evaluate a new method for creating random pattern/target pairs. This method
generates both satisfiable and unsatisfiable instances, and can produce computationally challenging
instances with only a few tens of vertices in the pattern, and 150 vertices in the target. Our work
builds upon the phase transition phenomena observed in satisfiability and graph colouring problems
first described by \citet{Cheeseman:1991} and \citet{Mitchell:1992}---?? cite later stuff and
explain really hard.

For subgraph isomorphism we identify three relevant control parameters: we can independently alter
the edge probability of the pattern graph, the edge probability of the target graph, and the
relative orders (number of vertices) of the pattern and target graphs.  For non-induced
isomorphisms, with the correct choice of parameters we see results very similar to those observed
with boolean satisfiability problems: there is a phase transition from satisfiable to unsatisfiable
(and we can predict its location), and we see a complexity peak occur with three different solvers
near this phase transition.

For certain choices of parameters for induced isomorphisms, there are two phase transitions, going
from satisfiable to unsatisfiable, and then from unsatisfiable back to satisfiable. Again, when
going from satisfiable to unsatisfiable (from either direction), instances go from being trivial to
really hard to solve. However, each of the three solvers we tested also finds the central
unsatisfiable region to be hard---this may be unexpected, since this region is not near a
phase transition. To show that this is not a simple weakness of current subgraph isomorphism
algorithms, we verify that this region is also hard when using a pseudo-boolean encoding, and
under reduction to the clique problem. Interestingly, the constrainedness measure
\citep{Gent:1996:Kappa} \emph{does} predict this difficult region---these instances provide (?? the
first??) evidence in favour of constrainedness, rather than proximity to a phase transition, being
an accurate predictor of difficulty.

@@

Our experiments were performed on systems with Intel Xeon E5-4650 v2 CPUs and 768GBytes RAM
(although this much RAM was not needed), running Scientific Linux release 6.7.   ?? Software
versions: Clasp 3.1.3, Gurobi 6.0.5, Glasgow, LAD version 2, VFLib. Software was compiled using GCC
4.9.

?? Stuff about runtimes and recursive calls. We are not aiming to compare solvers; rather, we are
looking for solver-independent behaviour. We need a quick note on the models used here too.

?? Timeouts

@@

Suppose we arbitrarily decide upon a pattern graph order of 20, a target graph order of 150, and a
fixed target edge probability of 0.40. As we vary the pattern edge probability from 0 to 1, we would
expect to see a shift from entirely satisfiable instances (with no edges in the pattern, we can
always find a match) to entirely unsatisfiable instances (a maximum clique in this order and edge
probability of target graph will usually have between 9 and 12 vertices). Indeed,
\cref{figure:phase-transition} shows that this is the case: the line (and the points show a subset
of the samples) ?? describe this a bit more, find the 0.5 SAT mark.

?? Some stuff about the complexity peak in the Glasgow solver. This looks remarkably similar to random
3SAT problems---compare, for example, Figure 1 of \citet{LeytonBrown:2014}. In particular,
satisfiable instances tend to be easier but show greater variation than unsatisfiable instances, and
there are exceptionally hard satisfiable instances \citep{Smith:1997}.

?? Are unsatisfiable instances on the phase transition much harder than satisfiable ones?

@@

What if we alter the edge probabilities for both the pattern graph and the target graph?  In the top
row of \cref{figure:non-induced} we show the satisfiability phase transition for the non-induced
variant, for patterns of order 10, 20 and 30, a target of order 150, and varying pattern (x-axis)
and target (y-axis) edge probabilities. Each axis runs over 51 edge probabilities, from 0 to 1 in
steps of 0.05; for each of these 2601 points, we generate ten random instances. The colour denotes
(?? wrong word?) the proportion of these instances which were found to be satisfiable.  Inside the
orange region, at the bottom right of each plot, every instance is unsatisfiable---here we are
trying to find a dense pattern in a sparse target. In the purple region, at the top left, every
instance is satisfiable---we are looking for a sparse pattern in a dense target (which is easy,
since we only have to preserve adjacency, not non-adjacency). The white band between the regions
shows the location of the phase transition: here, roughly half the instances are satisfiable. (We
discuss the black line below.)

@@

To predict the location of the phase transition, we can calculate the expected number of solutions
by making a few dubious assumptions involving rounding and independence. Since we are trying to find
an \emph{injective} mapping from a pattern $P = G(p, d_p)$ to a target $T = G(t, d_t)$, there are
$t^{\underline{p}} = t \cdot (t - 1) \cdot \ldots \cdot (t - p + 1)$ possible assignments of target
vertices to pattern vertices.  We expect the pattern to have $d_p \cdot \binom{p}{2}$ edges, so we
obtain the probability of all of these edges being mapped to edges in the target by raising $d_t$ to
this power, giving an expected number of solutions of \[ \langle Sol \rangle = t^{\underline{p}}
\cdot {d_t}^{d_p \cdot \binom{p}{2}} \textnormal{.} \] This formula predicts a very sharp phase
transition from $\langle Sol \rangle \ll 1$ to $\langle Sol \rangle \gg 1$. We plot where this
occurs using black lines in the first row of \cref{figure:non-induced}.

We can see that this prediction is generally accurate, except that for very low and very high
pattern densities, we overestimate the satisfiable region slightly. This is because although an
expected number of solutions much below one implies a high likelihood of unsatisfiability, it is not
true that a high expected number of solutions implies that any particular instance is likely to be
satisfiable---a similar behaviour is seen with random constraint satisfaction problems
\citep{Smith:1994,Smith:1996}.

?? Also because of half edges.

@@

When designing variable and value ordering heuristics for backtracking search algorithms, there are
various general principles which appear to be effective---one of these is to try to maximise the
expected number of solutions inside any subproblem considered during search \citep{Gent:1996:EN}.
This is usually done by cheaper surrogates, rather than direct calculation. When branching, both LAD
and Glasgow will pick a variable with fewest remaining values in its domain: doing this will
generally reduce the first part of the $\langle Sol \rangle$ equation by as little as possible. When
two or more domains are of equal size, LAD simply breaks ties lexicographically. Glasgow, however, will
pick a variable corresponding to a pattern vertex of highest degree. This strategy was determined
empirically, but could have been derived from the $\langle Sol \rangle$ formula: picking a
pattern vertex of high degree will make the remaining pattern subgraph sparser, which will decrease
the exponent in the second half of the formula, maximising the overall value. LAD does not apply a
value ordering heuristic, but Glasgow does: it prefers target vertices of lowest degree. Again, this
was determined empirically, but it has the effect of increasing $\langle Sol \rangle$ by increasing
the remaining target density.

@@

In \cref{figure:induced} we repeat our experiments, finding induced isomorphisms. With a pattern of
order 10, we get two independent phase transitions (?? describe properly). For each of the solvers,
the satisfiable instances before either phase transition are generally easy, instances near the
phase transition are hard, and the large central unsatisfiable region is again easy.

?? For larger patterns of order 20 and 30, we have large unsatisfiable regions in the middle,
but these remain hard.

Pattern orders 14, 15 and 16 show the transition between the two styles.

\subsection{Predicting Induced Behaviour}

By repeating the argument for locating the non-induced phase transition and additionally considering
non-edges, we get an expected number of solutions of \[ \langle Sol \rangle = t^{\underline{p}}
    \cdot {d_t}^{d_p \cdot \binom{p}{2}} \cdot {(1 - d_{t})}^{(1 - d_{p}) \cdot \binom{p}{2}}
\textnormal{.} \] We plot this
using black lines on \cref{figure:induced}---again, our prediction is accurate except for very
sparse or very dense patterns.

@@

\subsection{New Heuristics}

?? Explain why we might guess degree heuristics just won't work for induced: for everything we say
about degree, the opposite holds too for the complement constraints.

?? Possibly send the data to Lars, and get him to run his portfolios stuff on it, and then make fun
of machine learning?

\section{Other Encodings and Solvers}

@@

We create a pseudo-boolean (PB) encoding as follows. For each pattern vertex $v$ and each target
vertex $w$, we have a binary variable which takes the value 1 if and only if that $v$ is mapped to
$w$.  Constraints are added to ensure that each pattern vertex maps to exactly one target vertex,
that each target vertex is mapped to by at most one pattern vertex, that adjacent vertices are
mapped to adjacent vertices, and that non-adjacent vertices are mapped to non-adjacent vertices. We
used the Clasp solver~\citep{gekakaosscsc11a} to solve the pseudo-boolean instances. The instances
that are hard for the Glasgow solver remain hard for the PB solver, including instances inside the
central region, and the easy satisfiable instances remain easy.

>>>>

\author{First Author \thanks{This work was supported by a small pile of money and a large
    electricity bill.} \and Second Author
    \and Third Author \\
Some Instituion, Somewhere, Some Country \\
anonymous@example.com}

@@

    \caption{There is a non-induced isomorphism from the pattern graph, in the centre, to the target
    on the right, mapping vertex 1 to 1, 2 to 3, 3 to 5 and 4 to 6. This is not an induced
    isomorphism, since there is an edge between 1 and 5 in the target but not between 1 and 3 in the
    pattern. The mapping from the pattern to the (same) target on the left, sending 1 to 2, 2 to 6,
    3 to 5 and 4 to 3, is both non-induced and induced.}
    \label{figure:sip}
\end{figure}

@@

Here we present and evaluate a new method for creating random pattern/target pairs. This method
generates both satisfiable and unsatisfiable instances, and can produce computationally challenging
instances with only a few tens of vertices in the pattern, and 150 vertices in the target. Our work
builds upon the phase transition phenomena observed in satisfiability and graph colouring problems
first described by \citet{Cheeseman:1991} and \citet{Mitchell:1992}.  For subgraph isomorphism we
identify three relevant control parameters: we can independently alter the edge probability of the
pattern graph, the edge probability of the target graph, and the relative orders (number of
vertices) of the pattern and target graphs.  For non-induced isomorphisms, with the correct choice
of parameters we see results very similar to those observed with boolean satisfiability problems:
there is a phase transition from satisfiable to unsatisfiable (and we can predict its location), and
we see a complexity peak occur with three different solvers near this phase transition.

For certain choices of parameters for induced isomorphisms, there are two phase transitions, going
from satisfiable to unsatisfiable, and then from unsatisfiable back to satisfiable. Again, when
going from satisfiable to unsatisfiable (from either direction), instances go from being trivial to
really hard to solve. However, each of the three solvers we tested also finds the central
unsatisfiable region to be hard---this may be unexpected, since this region is not near a phase
transition. To show that this is not a simple weakness of current subgraph isomorphism algorithms,
we verify that this region is also hard when using a pseudo-boolean encoding, and under reduction to
the clique problem. Interestingly, the constrainedness measure \citep{Gent:1996:Kappa} \emph{does}
predict this difficult region---these instances provide evidence in favour of constrainedness,
rather than proximity to a phase transition, being an accurate predictor of difficulty.

@@

Our experiments were performed on systems with Intel Xeon E5-4650 v2 CPUs and 768GBytes RAM
(although this much RAM was not needed), running Scientific Linux release 6.7. We selected three
subgraph isomorphism solvers: the Glasgow solver \citep{McCreesh:2015}, LAD \citep{Solnon:2010}, and
VF2 \citep{Cordella:2004}; each was compiled using GCC 4.9. These solvers all use backtracking
search to build up an assignment of target vertices (values) to pattern vertices (variables), with
different kinds of inference and ordering heuristics.

In each case we measure the number of recursive calls (guessed assignments) made, not runtimes. We
are not aiming to compare absolute performance between solvers; rather, we are looking for
solver-independent patterns of difficulty. We used a timeout of 1,000 seconds, which was enough for
the Glasgow solver to solve nearly all our instances (whose orders were selected with this timeout
in mind), although we may slightly overestimate the proportion of unsatisfiable instances for
extremely sparse or dense pattern graphs. The LAD and VF2 solvers experienced many more failures
with this timeout, so our picture of just how hard the hardest instances are with these solvers is
less detailed.

@@

Suppose we arbitrarily decide upon a pattern graph order of 20, a target graph order of 150, and a
fixed target edge probability of 0.40. As we vary the pattern edge probability from 0 to 1, we would
expect to see a shift from entirely satisfiable instances (with no edges in the pattern, we can
always find a match) to entirely unsatisfiable instances (a maximum clique in this order and edge
probability of target graph will usually have between 9 and 12 vertices). The dark line in
\cref{figure:phase-transition} shows that this is the case. For densities of 0.67 or greater, no
instance is satisfiable; with densities of 0.44 or less, every instance is satisfiable; and with a
density of 0.55, roughly half the instances are satisfiable.

The light line plots mean search effort using the Glasgow solver: for sparse patterns, the problem
is trivial, for dense patterns proving unsatisfiability is not particularly difficult, and we see a
complexity peak around the point where half the instances are satisfiable.  We also plot the search
cost of individual instances, as points. The behaviour we observe looks remarkably similar to random
3SAT problems---compare, for example, Figure 1 of \citet{LeytonBrown:2014}. In particular,
satisfiable instances tend to be easier but show greater variation than unsatisfiable instances, and
there are exceptionally hard satisfiable instances \citep{Smith:1997}.

@@

What if we alter the edge probabilities for both the pattern graph and the target graph?  In the top
row of \cref{figure:non-induced} we show the satisfiability phase transition for the non-induced
variant, for patterns of order 10, 20 and 30, a target of order 150, and varying pattern (x-axis)
and target (y-axis) edge probabilities. Each axis runs over 51 edge probabilities, from 0 to 1 in
steps of 0.05; for each of these 2601 points, we generate ten random instances. The colour denotes
the proportion of these instances which were found to be satisfiable.  Inside the orange region, at
the bottom right of each plot, every instance is unsatisfiable---here we are trying to find a dense
pattern in a sparse target. In the purple region, at the top left, every instance is
satisfiable---we are looking for a sparse pattern in a dense target (which is easy, since we only
have to preserve adjacency, not non-adjacency). The white band between the regions shows the
location of the phase transition: here, roughly half the instances are satisfiable. (We discuss the
black line below.)

@@

We can approximately predict the location of the phase transition by calculating (with
simplifications regarding rounding and independence) the expected number of solutions for given
parameters. Since we are trying to find an \emph{injective} mapping from a pattern $P = G(p, d_p)$
to a target $T = G(t, d_t)$, there are $t^{\underline{p}} = t \cdot (t - 1) \cdot \ldots \cdot (t -
p + 1)$ possible assignments of target vertices to pattern vertices.  We expect the pattern to have
$d_p \cdot \binom{p}{2}$ edges, so we obtain the probability of all of these edges being mapped to
edges in the target by raising $d_t$ to this power, giving an expected number of solutions of \[
\langle Sol \rangle = t^{\underline{p}} \cdot {d_t}^{d_p \cdot \binom{p}{2}} \textnormal{.} \] This
formula predicts a very sharp phase transition from $\langle Sol \rangle \ll 1$ to $\langle Sol
\rangle \gg 1$. We plot where this occurs using black lines in the first row of
\cref{figure:non-induced}.

This prediction is generally reasonably accurate, except that for very low and very high pattern
densities, we overestimate the satisfiable region. This is due to variance: although an expected
number of solutions much below one implies a high likelihood of unsatisfiability, it is not true
that a high expected number of solutions implies that any particular instance is likely to be
satisfiable---a similar behaviour is seen with random constraint satisfaction problems
\citep{Smith:1994,Smith:1996}.

@@

When designing variable and value ordering heuristics for backtracking search algorithms, there are
various general principles which appear to be effective---one of these is to try to maximise the
expected number of solutions inside any subproblem considered during search \citep{Gent:1996:EN}.
This is usually done by cheaper surrogates, rather than direct calculation. When branching, both LAD
and Glasgow will pick a variable with fewest remaining values in its domain: doing this will
generally reduce the first part of the $\langle Sol \rangle$ equation by as little as possible. When
two or more domains are of equal size, LAD simply breaks ties lexicographically. Glasgow, however, will
pick a variable corresponding to a pattern vertex of highest degree. This strategy was apparently
determined empirically, but could have been derived from the $\langle Sol \rangle$ formula: picking
a pattern vertex of high degree will make the remaining pattern subgraph sparser, which will
decrease the exponent in the second half of the formula, maximising the overall value. LAD does not
apply a value ordering heuristic, but Glasgow does: it prefers target vertices of lowest degree.
Again, this appears to have been determined empirically, but it has the effect of increasing
$\langle Sol \rangle$ by increasing the remaining target density.

@@

In the first four rows of \cref{figure:induced} we repeat our experiments, finding induced
isomorphisms. With a pattern of order 10, we get two independent phase transitions: the bottom right
half of the plots resemble the non-induced results, and the top left half is close to a mirror
image. The central satisfiable region, which is away from either phase transition, is
computationally easy, but instances near the phase transition are hard.

For larger patterns of order 20 and 30, we have a large unsatisfiable region in the middle. Despite
not being near either phase transition, instances in the centre remain computationally challenging.
We also plot patterns of orders 14, 15 and 16, to show the transition between the two behaviours.

\subsection{Predictions and Heuristics}

For small patterns, simply considering the mirror image of the non-induced prediction provides a
reasonable approximation of the location of the phase transition, but as we show on the fifth row of
the figure, this breaks down for larger patterns. For a better prediction, we repeat the argument
for locating the non-induced phase transition and additionally considering non-edges, to get an
expected number of solutions of \[ \langle Sol \rangle = t^{\underline{p}} \cdot {d_t}^{d_p \cdot
\binom{p}{2}} \cdot {(1 - d_{t})}^{(1 - d_{p}) \cdot \binom{p}{2}} \textnormal{.} \] We plot
this using black lines on \cref{figure:induced}---again, our prediction is accurate except for
very sparse or very dense patterns.

We might guess that degree-based heuristics would just not work for the induced problem: for any
claim about the degree, the opposite will hold for the complement constraints. However, empirically,
this is not the case: on the final row of \cref{figure:induced}, we show whether it is better to use
the original pattern and target as the input to the Glasgow algorithm, or to take the complements.
(The only steps performed by the Glagsow algorithm which differs under taking the complements are
the degree-based heuristics.  LAD and VF2 are not symmetric in this way: LAD performs a filtering
step using degree information, but does not consider the complement degree, and VF2 uses
connectivity in the pattern graph.)

@@

@@

We create a pseudo-boolean (PB) encoding as follows. For each pattern vertex $v$ and each target
vertex $w$, we have a binary variable which takes the value 1 if and only if that $v$ is mapped to
$w$.  Constraints are added to ensure that each pattern vertex maps to exactly one target vertex,
that each target vertex is mapped to by at most one pattern vertex, that adjacent vertices are
mapped to adjacent vertices, and that non-adjacent vertices are mapped to non-adjacent vertices. We
used the Clasp solver~\citep{gekakaosscsc11a} version 3.1.3 to solve the pseudo-boolean instances.
The instances that are hard for the Glasgow solver remain hard for the PB solver, including
instances inside the central region, and the easy satisfiable instances remain easy.
